Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-09-04 paper/main.tex
whitespace in paper and update noteobok
Updates from Overleaf

\shorttitle{RR Lyrae stars in the Pal 5 stream}
\shortauthors{Price-Whelan et al.}
\newcommand{\sa}[1]{{\color{teal} SP: #1}}

@@

\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.

@@

\subsection{RRLs associated with Pal 5 and its tails}
We find a total of \NRRL\ RRLs with membership probability $> 0.5$ that lie within the stream track \citep[again using the track from][see lines in Figure~\ref{fig:members}]{Bonaca:2019}.
Of those, \NRRab\ are classified as type \typeab\ and \NRRc\ are classified as type \typec\ RRLs.
We find a total of \NRRcl\ RRLs within the estimated Jacobi radius of the cluster ($\approx 11'$), meaning that \NRRtails\ RRLs are spread between the leading and trailing tails.
\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.
No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.

@@

Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
This is closer to the sun than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.

@@

% \todo{APW: still typical?} 
%CM: yes, Nab/Nc has a long tail towards large values 
Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$,a typical ratio found Oosterhoff I populations \citep{Smith:1995}. 
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}. Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself. 
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field membership probabilities $>0.1$ (small markers) inside the Pal~5 track (large markers), colored by membership probability. The distribution of the RRLs found in the cluster and it's tails is consistent with an Oosterhoff type I (dashed line). }
\label{fig:PA_diagram}
\end{center}
\end{figure*}

>>>>

\shorttitle{RR Lyrae stars in the Pal 5 stream}
\shortauthors{Price-Whelan et al.}
\newcommand{\sa}[1]{{\color{teal} SP: #1}}
\newcommand{\ab}[1]{{\color{magenta} AB: #1}}

@@

\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, $d$, and proper motion components.
\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.

@@

\subsection{RRLs associated with Pal 5 and its tails}
We find a total of \NRRL\ RRLs with membership probability $> 0.5$ that lie within the stream track \citep[again using the track from][see lines in Figure~\ref{fig:members}]{Bonaca:2019}.
Of those, \NRRab\ are classified as type \typeab\ and \NRRc\ are classified as type \typec\ RRLs.
We find a total of \NRRcl\ RRLs within the estimated Jacobi radius of the cluster ($\approx 11'$), meaning that \NRRtails\ RRLs are spread between the leading and trailing tails.
\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.
No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
\ab{mention RRL overdensities along the tails don't correspond to MS overdensities?}

@@

\ab{this section is mostly about the distances -- include in the subtitle?}
Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
This is closer to the sun than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
\ab{can we work this point a bit more? I thought it was mainly the HST CMD measurement from Aaron that was an outlier, and everything else at a larger distance. then I think in Nitya + Tobias paper they found that to match pms, they require a smaller distance, and now we find it in PS1 RRL. I don't think we need to recount the whole history in a letter, but it might be helpful to comment on the difference wrt Kathy's RRL result}
We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.

@@

% \todo{APW: still typical?}
%CM: yes, Nab/Nc has a long tail towards large values
Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$,a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
\ab{can we provide a sentence for what it (physically) means to be an oosterhoff I population?}
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}. Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself.
\ab{this is interesting indeed -- can we provide a hypothesis?}
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field membership probabilities $>0.1$ (small markers) inside the Pal~5 track (large markers), colored by membership probability. The distribution of the RRLs found in the cluster and its tails is consistent with an Oosterhoff type I (dashed line). }
\label{fig:PA_diagram}
\end{center}
\end{figure*}
